---
title: "607HW10_Connin"
date: "4/13/21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Overview

This project recreates the code provided in Chapter 2 of Text Mining with R: A Tidy Approach. 

https://www.tidytextmining.com/sentiment.html


```{r}
#load libraries and get sentiment lexicon

library(tidytext)
library(textdata) #required to download lexicon
library(stringr)
library(dplyr)
library(janeaustenr)
library(wordcloud)
library(reshape2)

get_sentiments( "afinn")  # assigns relative value to word
get_sentiments( "nrc") # assigns category of emotion to word
get_sentiments( "bing") #assigns categories of positive or negative to word

```

### Create corpus

Import Jane Austen novels and convert text to corpus with tidy format. These texts did not require separate cleaning. 


Note: unnest_tokens -- creates one-token-per-row structure. A token is a meaningful unit of text (e.g., a word). This step cleans and tokenizes the text in a single step. 

```{r}

# Create corpus of Jane Austen novels. Create columns (identifiers) to track line number and chapter.   

tidy_books <- austen_books()%>%
  group_by(book)%>%
    mutate(linenumber = row_number(), 
    chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]", 
    ignore_case = TRUE)))) %>%
    ungroup()%>%
    unnest_tokens(word, text) # produces a tokenized word column; we will perform inner-joins and anti-joins on this 
```
### Match corpus for Emma and lexicon.

```{r}

# join intersection of nrc lexicon (positive words) and word tokens from book, Emma. 

nrc_joy <- get_sentiments("nrc") %>% 
  filter(sentiment == "joy") # grab the lexicon associated with joy words

tidy_books %>%
  filter(book == "Emma") %>%
  inner_join(nrc_joy) %>% #use inner_join to create the word matches
  count(word, sort = TRUE)  # counts the most common joy word associations in Emma

```
### Match corpus of all Austen books in package with lexicon

```{r}
library(tidyr)

jane_austen_sentiment <- tidy_books %>%
  inner_join(get_sentiments("bing")) %>%
  count(book, index = linenumber %/% 80, sentiment) %>% # count matches using integer division to break into 80 line sections
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>% #separate negative and positive counts
  mutate(sentiment = positive - negative)

```

### Plot sentiment scores

```{r}
library(ggplot2)

ggplot(jane_austen_sentiment, aes(index, sentiment, fill = book)) + # fill colorizes the graphs
  geom_col(show.legend = FALSE) +
  facet_wrap(~book, ncol = 2, scales = "free_x")
```
## Compare the three sentiment lexicons


```{r}

# Just use Pride and Prejudice

pride_prejudice <- tidy_books %>% 
  filter(book == "Pride & Prejudice")

afinn <- pride_prejudice %>% 
  inner_join(get_sentiments("afinn")) %>% 
  group_by(index = linenumber %/% 80) %>% 
  summarise(sentiment = sum(value)) %>% 
  mutate(method = "AFINN")

bing_and_nrc <- bind_rows(
  pride_prejudice %>% 
    inner_join(get_sentiments("bing")) %>%
    mutate(method = "Bing et al."), pride_prejudice %>% 
    inner_join(get_sentiments("nrc") %>% 
    filter(sentiment %in% c("positive", "negative"))) %>%
    mutate(method = "NRC")) %>%
    
    count(method, index = linenumber %/% 80, sentiment) %>%
    pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>% 
    mutate(sentiment = positive - negative)

# bind net sentiments for each chunk of text 

bind_rows(afinn, bing_and_nrc) %>%
  ggplot(aes(index, sentiment, fill = method)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~method, ncol = 1, scales = "free_y")


```
### Get lexicon

```{r}
get_sentiments("nrc") %>% 
  filter(sentiment %in% c("positive", "negative")) %>% 
  count(sentiment)

get_sentiments("bing") %>% 
  count(sentiment)

```
### Plot word matches

```{r}
bing_word_counts <- tidy_books %>%
    inner_join(get_sentiments("bing")) %>%
    count(word, sentiment, sort = TRUE) %>%
    ungroup() 


    #pipe into ggplot
    
bing_word_counts %>%
  group_by(sentiment) %>%
  slice_max(n, n = 10) %>% 
  ungroup() %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, scales = "free_y") +
  labs(x = "Contribution to sentiment",
       y = NULL)



```
### Customize stop words

```{r}
custom_stop_words <- bind_rows(tibble(word = c("miss"),  
                                      lexicon = c("custom")), 
                               stop_words)

custom_stop_words
```

## Create Word clouds

```{r}
library(wordcloud)

tidy_books %>%
  anti_join(stop_words) %>%
  count(word) %>%
  with(wordcloud(word, n, max.words = 100))
```
### Customize clouds

```{r}
tidy_books %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE) %>%
  acast(word ~ sentiment, value.var = "n", fill = 0) %>%
  comparison.cloud(colors = c("gray20", "gray80"),
                   max.words = 100)
```
## Other options


```{r}

#split dataframe by chapter using regex

austen_chapters <- austen_books() %>%
  group_by(book) %>%
  unnest_tokens(chapter, text, token = "regex", 
                pattern = "Chapter|CHAPTER [\\dIVXLC]") %>%
  ungroup()

austen_chapters %>% 
  group_by(book) %>% 
  summarise(chapters = n())
```
### Normalize word counts

```{r}
bingnegative <- get_sentiments("bing") %>% 
  filter(sentiment == "negative")

wordcounts <- tidy_books %>%
  group_by(book, chapter) %>%
  summarize(words = n())

tidy_books %>%
  semi_join(bingnegative) %>% #note semijoin
  group_by(book, chapter) %>%
  summarize(negativewords = n()) %>%
  left_join(wordcounts, by = c("book", "chapter")) %>% #note leftjoin
  mutate(ratio = negativewords/words) %>%
  filter(chapter != 0) %>%
  slice_max(ratio, n = 1) %>% 
  ungroup()
```

